\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-06-12 short.tex
workshop cross submission

\newcommand{\system}{\textsc{DreamCoder}~}
\newcommand{\systemEnding}{\textsc{DreamCoder}}
\newcommand{\lowerBound}{\mathscr{L}}
\newcommand{\code}[1]{{\footnotesize\texttt{#1}}}
\newcommand{\scode}[1]{{\tiny\texttt{#1}}}
\newcommand{\codechar}[1]{{\footnotesize{\texttt{"#1"}}}}
% Use the following line for the initial blind version submitted for review:
\usepackage[nohyperref]{icml2018}

@@

The FlashFill system~\cite{gulwani2011automating} embedded in
Microsoft Excel solves problems such as these and is probably the best
known practical program-induction algorithm, but researchers in
programming languages and AI have had successes in many domains, such as handwriting recognition and
generation~\cite{lake2015human}, question
answering~\cite{johnson2017clevr} and robot motion
planning~\cite{devlin2017neural}.  These systems
work in different ways, but most hinge upon a carefully engineered
\textbf{Domain Specific Language (DSL)}.  %% This is especially true for
%% systems like FlashFill that induce a wide range of programs
%% very quickly, in a few seconds or less.  
DSLs constrain the search
over programs with strong prior knowledge in the form of a restricted
library of primitives tuned to the domain: for
text editing, these are operations like appending and
splitting on strings.

@@

We take inspiration from several ways that skilled human
programmers have learned to code: Skilled coders build libraries of
reusable subroutines that are shared across related programming tasks,
and can be composed to generate increasingly complex
subroutines.  In text editing, a good library should support routines
for splitting on characters, but also specialize these routines to
split on frequent delimiters such as spaces or commas. Skilled coders
also learn to recognize what kinds of programming idioms and library
routines would be useful for solving the task at hand, even if they
cannot instantly work out the details.  In text editing, one might
learn that if outputs are consistently shorter than inputs, removing
characters is likely part of the solution.%% ; if every output
%% contains a constant substring (e.g., ``Dr.''), inserting or appending
%% that constant string is likely to be a subroutine.

Our algorithm is called \system because it incorporates these insights into a
novel kind of ``wake-sleep'' learning (c.f. \cite{hinton1995wake}), iterating
between ``wake'' and ``sleep'' phases to achieve three goals: finding
programs that solve tasks; creating a DSL by discovering
and reusing domain-specific subroutines; and training a neural network
that efficiently guides search for programs in the DSL.
The learned DSL effectively encodes a prior on programs likely to
solve tasks in the domain, while the neural net looks at the example
input-output pairs for a specific task and produces a ``posterior''
for programs likely to solve that specific task.  The neural network
thus functions as a \textbf{recognition model} supporting a form of
approximate Bayesian program induction, jointly trained with a
\textbf{generative model} for programs encoded in the DSL, in the
spirit of the Helmholtz machine~\cite{hinton1995wake}. The
recognition model ensures that searching for programs remains
tractable even as the DSL (and hence the search space for programs)
expands.

%% Our \system
%% algorithm incorporates these insights by iterating
Concretely, our algorithm iterates through three cycles:
\\\noindent\textbf{Wake Cycle}: Given a set of \textbf{tasks}, typically
several hundred, we search for compact programs that solve these tasks.
Currently we take the simple strategy of
enumerating programs written in the current DSL in order of their probability according to the neural network.
\\\noindent\textbf{Sleep-G Cycle}: Here we grow the the DSL (or \textbf{G}enerative model),
which allows the agent to more compactly
write programs in the domain. We modify the structure of the DSL by
discovering regularities across programs found during waking, compressing
them to distill out common code fragments across successful programs.
The technical machinery behind this DSL learning
comes from a formalism known as Fragment Grammars~\cite{tim},
which casts DSL induction as grammar induction.
\\\noindent\textbf{Sleep-R Cycle}: This trains a neural net (the \textbf{R}ecognition model) to
predict a distribution over programs written in the current DSL, in the spirit of ``amortized'' or
``compiled'' inference~\cite{le2016inference}.
The network is trained, conditioned on a task, to assign high probability to
programs that have high prior probability according to the current DSL,
while also assigning high likelihood to the task at hand,
thus amortizing the cost of finding programs with high posterior probability.

@@

We apply \system to list processing %% (Section~\ref{listSection})
and text editing, %% (Section~\ref{textSection})
using a recurrent network for
the recognition model, and initially providing the system with generic sequence manipulation primitives:
\code{foldr}, \code{unfold}, \code{if}, \code{map}, \code{length},
\code{index}, \code{=}, \code{+}, \code{-}, \code{0}, \code{1}, \code{cons},
\code{car}, \code{cdr}, \code{nil}, and \code{is-nil}.

@@

\textbf{Symbolic Regression: Programs from visual input.}
We apply \system
to symbolic regression problems.  Here, the
agent observes points along the curve of a function, and must write a
program that fits those points.  We initially equip our learner with
addition, multiplication, and division, and task it with solving
100 %% symbolic regression
problems, each either a polynomial or rational function.  The recognition model is a
convnet that observes an image of the target function's
graph (Fig.~\ref{functions}) --- visually, different kinds of
polynomials and rational functions produce different kinds of graphs,
and so the convnet can  look at a graph and predict
what kind of function best explains it.
%% These programs can contain real numbers,
%% which we set with gradient descent,
%% and penalize continuous parameters by
%% incorporating a BIC penalty into the likelihood model $\probability[x|p]$.
%% A key difficulty, however, is
%% that these problems are best solved with programs containing real
%% numbers.  Our solution to this difficulty is to enumerate
%%  programs with real-valued parameters, and then fit those
%% parameters by automatically differentiating through the programs the
%% system writes and use gradient descent to fit the parameters.
%% We define the likelihood model, $\probability[x|p]$, by assuming a Gaussian noise model for the input/output examples,
%% and penalize the use of real-valued parameters using the BIC~\cite{Bishop:2006:PRM:1162264}.
\begin{figure}\vspace{-0.0cm} \newcommand{\functionSize}{1.1cm}\centering
  \begin{minipage}[c]{0.3\columnwidth}%\vspace{-0.3cm}
  \includegraphics[width = \functionSize]{figures/functions/6.png}
  \includegraphics[width = \functionSize]{figures/functions/48.png}\\
  \includegraphics[width = \functionSize]{figures/functions/102.png}
  \includegraphics[width = \functionSize]{figures/functions/116.png}\\
  \vspace{1pt}%\includegraphics[width = \functionSize]{figures/functions/181.png}
  %\includegraphics[width = \functionSize]{figures/functions/160.png}
  \includegraphics[width = \functionSize]{figures/functions/160.png}
    \includegraphics[width = \functionSize]{figures/functions/149.png}
  \end{minipage}
  \begin{minipage}[c]{0.69\columnwidth}    
    \caption{Recognition model input for symbolic regression. DSL learns subroutines for polynomials (top rows) and rational functions (bottom) while the recognition  model jointly learns to look at a graph of the function (left) and predict which learned subroutines best explain the observation.}\label{functions}\vspace{-1cm}
        \end{minipage}
\end{figure}
%% We learn a DSL containing 13 new functions,
%% mainly templates for polynomials of different orders or ratios of polynomials.
%% The model also learns to find programs minimizing the number of continuous parameters ---
%% learning to represent linear functions with 
%% \code{(* real (+ x real))}, which has two continuous parameters, and represents quartic functions using $f_4$ in the rightmost column of Fig.~\ref{initialExampleDSL}
%% which has five continuous parameters.
%% This phenomenon arises from our Bayesian framing.

@@

\begin{figure*}\centering
  \includegraphics[width = 4.5cm]{figures/listLearningCurve.eps} \qquad
  \includegraphics[width = 4.5cm]{figures/textLearningCurve.eps}\qquad
  \includegraphics[width = 4.5cm]{figures/rationalCurve.eps}
\vspace{-0.4cm}  \caption{Learning curves for \system both with (\orange{in orange}) and without
    (\teal{in teal}) the recognition model. Iterations: \# wake/sleep cycles. Solid lines: \% holdout testing tasks solved w/ 10m timeout. Dashed lines: Average solve time.}\label{learningCurves}\vspace{-0.5cm}
\end{figure*}

@@

%\paragraph{Outlook.}
We contribute an algorithm, \systemEnding, that learns to program by
bootstrapping a DSL with new domain-specific primitives that the algorithm
itself discovers, together with a neural recognition model that learns to
efficiently deploy the DSL on new tasks. %% We believe this integration of top-down
%% symbolic representations and bottom-up neural nets --- both of them learned
%% --- can help make program induction systems more generally useful for AI. 
%Many
%directions remain open.
%\paragraph{Future.}
Two immediate future goals are to integrate more sophisticated neural recognition
models~\cite{devlin2017robustfill} and program
synthesizers~\cite{solar2008program}, which may improve performance in some
domains over the generic methods used here.
Another direction is DSL meta-learning: can we find a
\emph{single} universal primitive set that could bootstrap DSLs for
new domains, including the domains considered here,  but also many others?

>>>>

\newcommand{\system}{\textsc{DreamCoder}~}
\newcommand{\systemEnding}{\textsc{DreamCoder}}
\newcommand{\lowerBound}{\mathscr{L}}
\newcommand{\code}[1]{{\footnotesize\texttt{#1}}}
\newcommand{\scode}[1]{{\tiny\texttt{#1}}}
\newcommand{\codechar}[1]{{\footnotesize{\texttt{"#1"}}}}
% Use the following line for the initial blind version submitted for review:
\usepackage[nohyperref,accepted]{icml2018}

@@

The FlashFill system~\cite{gulwani2011automating} embedded in
Microsoft Excel solves problems such as these and is probably the best
known practical program-induction algorithm, but researchers %%  in
%% programming languages and AI 
have had many successes: handwriting recognition and
generation~\cite{lake2015human}; question
answering~\cite{johnson2017clevr}; and robot motion
planning~\cite{devlin2017neural}.  These systems
work in different ways, but most hinge upon a carefully engineered
\textbf{Domain Specific Language (DSL)}.  %% This is especially true for
%% systems like FlashFill that induce a wide range of programs
%% very quickly, in a few seconds or less.  
DSLs constrain the space
of programs with strong prior knowledge in the form of a restricted
set of primitives tuned to the domain: for
text editing, these are operations like appending and
splitting on characters.

@@

We are inspired by several ways that skilled human
programmers have learned to code: Skilled coders build libraries of
reusable subroutines shared across related programming tasks,
that can be composed to generate increasingly complex
subroutines.  In text editing, a good library should support routines
for splitting on characters, but also specialize these routines to
split on frequent delimiters like spaces. Skilled coders
also learn to recognize which programming idioms and library
routines are useful for solving the task at hand, even if they
cannot instantly work out the details.  %% In text editing, one might
%% learn that if outputs are consistently shorter than inputs, removing
%% characters is likely part of the solution.
%% ; if every output
%% contains a constant substring (e.g., ``Dr.''), inserting or appending
%% that constant string is likely to be a subroutine.

Our algorithm is called \system because it incorporates these insights into a
novel kind of ``wake-sleep'' learning (c.f. \citep{hinton1995wake}), iterating
between ``wake'' and ``sleep'' phases to achieve three goals: finding
programs that solve tasks; growing a DSL by discovering
and reusing domain-specific subroutines; and training a neural network
that  guides search for programs in the DSL.
The learned DSL effectively encodes a prior on programs likely to
solve tasks in the domain, while the neural net looks at %the example
%input-output pairs for
a specific task and produces a ``posterior''
for programs likely to solve that specific task.  The neural network
thus functions as a \textbf{recognition model} supporting a form of
approximate Bayesian program induction, jointly trained with a
\textbf{generative model} for programs encoded in the DSL, in the
spirit of the Helmholtz machine~\cite{hinton1995wake}. The
recognition model ensures that searching for programs remains
tractable even as the DSL 
expands.

%% Our \system
%% algorithm incorporates these insights by iterating
Concretely, our algorithm iterates through three cycles:
\\\noindent\textbf{Wake Cycle}: Given a set of \textbf{tasks}, 
 we search for compact programs that solve these tasks,
taking the simple strategy of
enumerating programs written in the current DSL in order of their probability according to the neural network.
\\\noindent\textbf{Sleep-G Cycle}: Here we grow  the DSL (\textbf{G}enerative model),
allowing the agent to more compactly
write programs in the domain. We modify the DSL by
discovering regularities across programs found during waking, compressing
them to distill out common code fragments. % across successful programs.
The technical machinery behind this DSL learning
comes from a formalism known as Fragment Grammars~\cite{tim}.
%which casts DSL induction as grammar induction.
\\\noindent\textbf{Sleep-R Cycle}: This trains a neural net (\textbf{R}ecognition model) to
predict a distribution over programs in the current DSL, in the spirit of ``amortized''  inference~\cite{le2016inference}.
The network is trained, conditioned on a task, to assign high probability to
programs that have high prior probability according to the DSL,
while also assigning high likelihood to the task at hand,
thus amortizing the cost of finding programs with high posterior probability.

@@

We apply \system to list processing %% (Section~\ref{listSection})
and text editing, %% (Section~\ref{textSection})
using a recurrent network for
the recognition model, initially providing the system with generic sequence manipulation primitives:
\code{foldr}, \code{unfold}, \code{if}, \code{map}, \code{length},
\code{index}, \code{=}, \code{+}, \code{-}, \code{0}, \code{1}, \code{cons},
\code{car}, \code{cdr}, \code{nil}, and \code{is-nil}.

@@

\textbf{Symbolic Regression: Programs from visual input.}
We apply \system
to symbolic regression problems.  Here, the
agent observes points along the curve of a function, and must write a
program that fits those points.  We initially equip our learner with
addition, multiplication, and division, and task it with solving
100 %% symbolic regression
problems, each either a polynomial or rational function.  The recognition model is a
convnet that observes an image of the target function's
graph (Fig.~\ref{functions}) --- visually, different kinds of
polynomials and rational functions produce different kinds of graphs,
and so the convnet can  look at a graph and predict
what kind of function best explains it.
%% These programs can contain real numbers,
%% which we set with gradient descent,
%% and penalize continuous parameters by
%% incorporating a BIC penalty into the likelihood model $\probability[x|p]$.
%% A key difficulty, however, is
%% that these problems are best solved with programs containing real
%% numbers.  Our solution to this difficulty is to enumerate
%%  programs with real-valued parameters, and then fit those
%% parameters by automatically differentiating through the programs the
%% system writes and use gradient descent to fit the parameters.
%% We define the likelihood model, $\probability[x|p]$, by assuming a Gaussian noise model for the input/output examples,
%% and penalize the use of real-valued parameters using the BIC~\cite{Bishop:2006:PRM:1162264}.
\begin{figure}\vspace{-0.1cm} \newcommand{\functionSize}{1.1cm}\centering
  \begin{minipage}[c]{0.3\columnwidth}\vspace{0.5cm}
  \includegraphics[width = \functionSize]{figures/functions/6.png}
  \includegraphics[width = \functionSize]{figures/functions/48.png}\\
  \includegraphics[width = \functionSize]{figures/functions/102.png}
  \includegraphics[width = \functionSize]{figures/functions/116.png}\\
  \vspace{1pt}%\includegraphics[width = \functionSize]{figures/functions/181.png}
  %\includegraphics[width = \functionSize]{figures/functions/160.png}
  \includegraphics[width = \functionSize]{figures/functions/160.png}
    \includegraphics[width = \functionSize]{figures/functions/149.png}
  \end{minipage}
  \begin{minipage}[c]{0.69\columnwidth}    
    \caption{Recognition model input for symbolic regression. DSL learns subroutines for polynomials (top rows) and rational functions (bottom) while the recognition  model jointly learns to look at a graph of the function (left) and predict which learned subroutines best explain the observation.}\label{functions}\vspace{-1cm}
        \end{minipage}
\end{figure}
%% We learn a DSL containing 13 new functions,
%% mainly templates for polynomials of different orders or ratios of polynomials.
%% The model also learns to find programs minimizing the number of continuous parameters ---
%% learning to represent linear functions with 
%% \code{(* real (+ x real))}, which has two continuous parameters, and represents quartic functions using $f_4$ in the rightmost column of Fig.~\ref{initialExampleDSL}
%% which has five continuous parameters.
%% This phenomenon arises from our Bayesian framing.

@@

\begin{figure*}\centering
  \includegraphics[width = 4.5cm]{figures/listLearningCurve.eps} \qquad
  \includegraphics[width = 4.5cm]{figures/textLearningCurve.eps}\qquad
  \includegraphics[width = 4.5cm]{figures/rationalCurve.eps}
  %\vspace{-0.4cm}
  \caption{Learning curves for \system both with (\orange{in orange}) and without
    (\teal{in teal}) the recognition model. Iterations: \# wake/sleep cycles. Solid lines: \% holdout testing tasks solved w/ 10m timeout. Dashed lines: Average solve time.}\label{learningCurves}\vspace{-0.5cm}
\end{figure*}

@@

%\paragraph{Outlook.}
We contribute an algorithm, \systemEnding, that learns to program by
bootstrapping a DSL with new domain-specific primitives that the algorithm
itself discovers, together with a neural recognition model that learns to
efficiently deploy the DSL on new tasks. %% We believe this integration of top-down
%% symbolic representations and bottom-up neural nets --- both of them learned
%% --- can help make program induction systems more generally useful for AI. 
%Many
%directions remain open.
%\paragraph{Future.}
%% Two immediate future goals are to integrate more sophisticated neural recognition
%% models~\cite{devlin2017robustfill} and program
%% synthesizers~\cite{solar2008program}, which may improve performance in some
%% domains over the generic methods used here.
One future direction is DSL meta-learning: can we find a
\emph{single} universal primitive set that could bootstrap DSLs for
new domains, including the domains considered here,  but also many others?
